Theory of surface second-harmonic generation for semiconductors including effects of nonlocal operators

### 2014-12-22 corrections/corrections.tex
preparing corrections by valerie and nicolas

\documentclass[floatfix,prb,aps,superscriptaddress,showpacs,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%\usepackage{showlabels}

\pacs{78.68.+m, 42.65.An, 71.15.Mb, 42.65.Ky, 78.66.-w}

\begin{document}

\title{Theory of surface second-harmonic generation for 
       semiconductors including effects of nonlocal operators}

\author{Sean M. Anderson}
    \affiliation{Centro de Investigaciones en Optica, 
                Le\'on, Guanajuato, M\'exico}
\author{Nicolas Tancogne-Dejean}
    \affiliation{Laboratoire des Solides Irradi\'es, \'Ecole Polytechnique, 
                CNRS, CEA/DSM, 91128 Palaiseau, France}
    \affiliation{European Theoretical Spectroscopy Facility (ETSF), 
                Palaiseau, France}
\author{Bernardo S. Mendoza}\email{bms@cio.mx}
    \affiliation{Centro de Investigaciones en Optica, 
                Le\'on, Guanajuato, M\'exico}
\author{Val\'erie V\'eniard}
    \affiliation{Laboratoire des Solides Irradi\'es, \'Ecole Polytechnique, 
                CNRS, CEA/DSM, 91128 Palaiseau, France}
    \affiliation{European Theoretical Spectroscopy Facility (ETSF), 
                Palaiseau, France}

\begin{abstract}
We formulate a theoretical approach of surface second-harmonic generation 
from semiconductor surfaces based on the length gauge and the 
electron  
density operator.  
Within the independent particle approximation 
the nonlinear second-order surface susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$    
is calculated, including for the first time in one unique formulation
(i) the scissors correction, needed to have the correct value of the
energy band gap,
(ii) the contribution of the nonlocal part of the  
pseudopotentials, routinely  used in \textit{ab initio} band structure calculations, 
and (iii) the derivation for the inclusion of the cut function, 
used to extract the surface response.
The first two contributions
are described by spatially nonlocal quantum
mechanical operators and are fully taken into account in the present formulation. 
As a test-case of the approach we calculate 
$\chi^{xxx}(-2\omega;\omega,\omega)$ for the clean Si(001)$2\times 1$
reconstructed surface and find that our scheme to extract the surface
through the cut function
readily works and is exact. The effects of the scissors correction and
of the nonlocal part of the pseudopotentials are discussed 
in surface nonlinear optics. 
The scissors correction shifts the spectrum to 
higher energies though the shifting is not rigid and mixes the 
$1\omega$ and $2\omega$ resonances, 
and has a strong influence in the line-shape. 
The 
effects of the nonlocal part of the pseudopotentials 
keeps the same line-shape of $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$, but 
reduces its value
by 15-20\%.  
Therefore, the inclusion of the three aforementioned 
contributions is very important and 
makes our scheme unprecedented and opens the possibility to 
study surface second-harmonic generation
with more versatility and providing more accurate results.
\end{abstract}

\maketitle

\section{Introduction}\label{intro}

In recent years surface nonlinear optical spectroscopies, particulary surface 
second-harmonic generation (SSHG), have evolved as useful nondestructive 
and noninvasive tools to study surface and interface properties. These 
properties include atomic structure, phase transitions, adsorption of 
atoms, and many others.\cite{daumPRL93, mcgilpOE94, meyerPRL95, powerPRL95, 
godefroyAPL96, hoferAPA96, dadapPRB97, bloembergenAPB99, mcgilpSRL99, 
suzukiAPB99, mitchellSS01, hughesPRB96, guyotPRB88, downerPSSA01, shenAPB99, 
shenNAT89, chenPRL81, mendozaPRL98, downerSIA01} Nowadays, SSHG spectroscopy 
is a crucial tool for research and development in microelectronics,
\cite{zheltikovLP00} semiconductors, \cite{lupkeSSR99} nanomaterials,
\cite{salazar-aparicioPRB14} and many more recent areas of great scientific 
and commercial interest.\cite{cazzanelliNM14} The high surface sensitivity 
of SSHG spectroscopy is due to the fact that within the dipole approximation 
the bulk SHG signal of centrosymmetric materials is identically zero.
Radiation can occur only at the surface where the inversion symmetry is 
broken. 
The bulk quadrupole contribution for centrosymmetric materials is
different from zero, but usually it is very small,\cite{downerSIA01} and we neglect it. 
Much of the foundation of surface science has been built from 
experiments involving emission or scattering of electrons from surfaces. 
These require ultrahigh vacuum (UHV) environments and provide no access to 
buried interfaces. However, SSHG is compatibile with non-UHV conditions and 
has access to interfaces buried beneath transparent overlayers. Even when 
applied to surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

The usefulness of SSHG could be limited by the lack of microscopic theoretical
understanding of the nonlinear spectra. The macroscopic phenomenological 
theory of SSHG, which relates the intensity, phase, and polarization of 
detected fields to nonlinear and linear susceptibilities at the material 
interface is now fully developed.\cite{downerSIA01}
 However, microscopic theory that relates 
electronic-level structure to the nonlinear source polarization is still being 
developed. \cite{butcherPOPS63, aspnesPRB72, sipePRB93,
  levinePRB94,aversaPRB95, hughesPRB96, rashkeevPRB98,beyond}  
 Within the independent 
particle approximation (IPA) some frameworks for bulk SHG have been developed 
to study the nonlinear optical response of bulk materials. \cite{butcherPOPS63, aspnesPRB72, sipePRB93, levinePRB94,aversaPRB95, hughesPRB96, rashkeevPRB98} 
In this article we put forward an approach to calculate the
microscopic second-harmonic surface susceptibility that encompasses several
theoretical features not taken into account before 
in the case of a surface.

The most used framework for \textit{ab initio} calculations, Density 
Functional Theory (DFT) within the Local Density Approximation (LDA),
\cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
well understood that one has to include the many-body interaction to correct 
for this underestimation of the gap. In this context, the so-called GW 
approximation\cite{onidaRMP02} is known to correct the electronic gap of most 
semiconductors\cite{luceroJPCM12}. However this can be a very expensive
calculation and thus one uses the much simpler scissors operator scheme.
\cite{levinePRL89,levinePRL91,delsolePRB93}  
This allows us to ``open'' the DFT-LDA gap to 
its correct experimental or GW value even when experiments are not available. 
In this context, to correct for the underestimation of the energy band gap of 
semiconductors Nastos et al.\cite{nastosPRB05} used the ``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to correctly include the 
many-body corrections through the scissors operator in the SH susceptibility.
Later, Cabellos \textit{et al}.\cite{cabellosPRB09} elaborated a derivation 
of the ``velocity gauge'' or ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the 
scissors operator and proved gauge invariance with respect to the length 
gauge. From these works it is clear the length gauge is a much better starting
point to obtain the surface SH (SSH) susceptibility, as will be elaborated 
in this article. However, these considerations are only valid for bulk 
semiconductors.

Concerning the optical response of surfaces and interfaces, Reining 
\textit{et al.}\cite{reiningPRB94} introduced the concept of a cut function 
in order to obtain the surface SH susceptibility tensor. This cut function 
is required since one usually uses a slab approach when treating semi-infinite 
surface systems.\cite{reiningPRB94} If the slab is centrosymmetric the susceptibility tensor will 
be identically zero. The cut function is such that it separates the nonlinear 
response for the two surfaces of the slab avoiding the destructive 
interference between them giving a finite value that one identifies with the 
SSH susceptibility tensor. If the slab is not centrosymmetric the cut function 
can be used to separate the different signals coming from either surface of 
the slab. Indeed, one of the main results of this article is to show that the 
SSH  susceptibility tensor obtained by using the cut function is correctly 
extracted from the slab. After Reining \textit{et al.},\cite{reiningPRB94}
Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02,sanoPRB02}
followed 
upon this work and in particular Ref.~\onlinecite{arzatePRB01} went into a 
detailed analysis of the different contributions to the SHG spectra of a 
surface and the nuanced relationship between bulk, surface, interband, 
intraband, $1\omega$ and $2\omega$ terms, and
Ref. \onlinecite{mejiaRMF04} developed a layer-by-layer 
analysis for the nonlinear responses of semiconductor systems, within a 
tight-binding framework. This model allows for obtaining results from 
selected regions of a system including the surface. However, in these 
references the scissors operator is either excluded or incorrectly 
implemented.

Finally, DFT-LDA calculations are often based on the use of pseudopotentials.
As it will be discussed 
in this article, the presence of a nonlocal part of the
pseudopotential introduces
corrections to
the momentum operator of the electron that have to be included with care in 
the SSH susceptibility. 
For the bulk counterpart see for instance Refs.~\onlinecite{ismailPRL01,luppiPRB08}.

Therefore, within the IPA the most complete approach for the calculation of 
the SSH susceptibility is one which includes (i) the scissors correction, 
(ii) the contribution of the nonlocal part of the 
pseudopotential, and
(iii) the cut function.
 Therefore, the goal of this paper is to derive a new 
expression within the length gauge for the SSH susceptibility tensor 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ 
that includes the aforementioned contributions. The inclusion of these three 
contributions makes our scheme unprecedented and opens the possibility to 
study surface SHG with more versatility and providing accurate results.

The paper is organized as follows. In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
length gauge formalism. This derivation includes the addition of the terms 
mentioned above that have been absent in previous works. Also, our 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ can be used 
for a layer-by-layer analysis if desired. 
Nevertheless, such an analysis could depend on the choice of the basis
used to expand the wavefunctions.
In Sec. \ref{results} we show the 
results of this reformulation with a study of a clean Si(001) surface with a 
$2 \times 1$ reconstruction, 
proving the correctness of our approach with a special test-case,
never exploited before.
We compare results from before and after adding 
the different nonlocal contributions. Finally, in Sec. \ref{conc} we give 
our conclusions.

\section{Theory}

\label{theory}

In this section we present the scheme used to calculate the surface 
second-order nonlinear response using the length gauge formalism and
the electron 
density operator. 
Some of the results presented  in Subsections \ref{pa} and \ref{lgf}
have already been discussed in earlier
studies and can be applied to bulk and surface studies.\cite{sipePRB00,cabellosPRB09}
We present them in order to have a self-contained derivation.
The terms presented in these Subsections are used in Subsection \ref{cd}, 
where we derive the expression 
for the surface $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$.

\subsection{Perturbation approach}\label{pa}

We assume the IPA, a classical electromagnetic field, and quantum mechanical 
matter. We neglect local field and excitonic effects. We can describe the 
system using 
the one electron density operator ${\rho}$, with which we can calculate the 
expectation value of a single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle = \mbox{Tr}({\rho}{\cal O}),\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr the 
trace. The density operator satisfies
$i\hbar(d{\rho}/dt) = [H(t), {\rho}], \label{rho}$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t) = H_{0} + H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the electron 
with the electromagnetic field.
To proceed with the solution of $\rho$ it is convenient to use the interaction 
picture, where a unitary operator $U = \exp({iH_{0}t/\hbar})$ transforms any 
operator $\mathcal{O}$ into $\tilde{\mathcal{O}} = U{\cal O}U^{\dagger}$. 
Even if $\mathcal{O}$ is time-independent, $\tilde{\mathcal{O}}$ is 
time-dependent through the explicit time dependence of $U$. The dynamical 
equation for $\tilde{\rho}$ is given by
\begin{equation*}
i\hbar \frac{d\tilde{\rho}}{dt} = [\tilde{H}_{I}(t), \tilde{\rho}],
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{\rho}(t) = 
    i\hbar \tilde{\rho}_{0} + 
    \int_{-\infty}^{t}dt^{\prime}
    [\tilde{H}_{I}(t^{\prime}),\tilde{\rho}(t^{\prime})],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0} = \tilde{\rho}(t = -\infty)$ is the unperturbed 
density matrix. We look for the standard perturbation series solution, 
$\tilde{\rho}(t) = \tilde{\rho}^{(0)} + \tilde{\rho}^{(1)} + 
\tilde{\rho}^{(2)} + \cdots$, where the superscript denotes the order (power) 
with which each term depends on the perturbation $H_{I}(t)$. From 
Eq.~\eqref{trans} the $N$-th order term is 
\begin{equation}
\tilde{\rho}^{(N)}(t) = 
    \frac{1}{i\hbar}\int_{-\infty }^{t}dt^{\prime}
    [\tilde{H}_{I}(t^{\prime }), \tilde{\rho}^{(N - 1)}(t^{\prime})].
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}^{(0)}
\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle = 
f(\hbar\omega_{n}(\mathbf{k}))\equiv f_{n}$. For a clean, cold semiconductor 
$f_{n}=1$ when $n$ is a valence ($v$) or occupied band, and zero when $n$ is 
a conduction ($c$) or empty band. 
We assume this for the remainder of the  
article.  
As we neglect spin-orbit coupling, the
final expression for
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$   
has to be multiplied by a factor of 2
to account for spin-degeneracy.
The expectation values must satisfy
$\langle{\cal O}\rangle = \mbox{Tr}({\rho}{\cal O}) = 
\mbox{Tr}(\tilde{\rho}\tilde{\cal O})$.

We will look for the expectation value of the microscopic current density, 
$\mathbf{J}$, given by 
\begin{equation*}
\mathbf{J} = \langle{\mathbf{J}}\rangle 
           = \frac{e}{A}\mbox{Tr}({\rho}\dot{\mathbf{r}}),
\end{equation*}
where $\dot{\mathbf{r}}$ is the time derivative of the position operator of the
electron with charge $e$, defined as
\begin{equation}
\mathbf{v}\equiv \dot{\mathbf{r}}=\frac{1}{i\hbar }[\mathbf{r},H_0],  
\label{mv}
\end{equation}
with $\mathbf{v}$ the velocity operator of the electron, and $A$ the
area of the unit cell. We calculate the polarization density 
$\mathbf{P}$, related to $\mathbf{J}$ by
$\mathbf{J}=d\mathbf{P}/dt$. For a 
perturbing electromagnetic field, $\mathbf{E}(t)=
\mathbf{E}(\omega )e^{-i\tilde{\omega} t}+c.c.$,
where $\tilde\omega=\omega+i\eta $,
and $\eta >0$ adiabatically switches on the interaction,
we write the second-order nonlinear
polarization as, 
\begin{equation}
{\mathbf{\cal P}}^{\mathrm{a}}(2\omega)=
\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)
E^{\mathrm{b}}(\omega)E^{\mathrm{c}}(\omega),  
\label{pshg}
\end{equation}
where $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega ;\omega ,\omega )$ is the nonlinear
susceptibility responsible for surface second-harmonic generation
(SSHG). 
The 
superscripts in Eq.~\eqref{pshg} denote Cartesian components, and if
repeated are to be summed over. Without loss of generality we will
define $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$
 to satisfy intrinsic permutation
symmetry, 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega ;\omega ,\omega )
=\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega ;\omega,\omega )$.

The unperturbed Hamiltonian 
is used to solve the Kohn-Sham equations\cite{kohnPR65} of Density  
Functional Theory (DFT). It is convenient to work within the Local 
Density Approximation (LDA), so we label the hamiltonian with the corresponding  
LDA superscript. Any other approximation can be used (like GGA) and our 
derivation remains the same. Then,
\begin{equation*}
H^{\mathrm{LDA}}_{0}(\mathbf{r},\mathbf{p})
=\frac{p^{2}}{2m_e}+V(\mathbf{r},\mathbf{p})
\end{equation*}
with $m_e$ the mass of the electron, $\mathbf{p}$ its canonical momentum, and 
$V$
 the periodic crystal potential, where we neglect spin-orbit terms.
To be more general in our derivation of
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$, we asume
contribution as is customary for most
pseudopotentials, and then we replace $V$ with
\begin{align*}
V^{\mathrm{ps}}(\mathbf{r},\mathbf{p})=V(\mathbf{r})+V^{\mathrm{nl}}(\mathbf{r},\mathbf{p})
,
\end{align*}
where 
$V(\mathbf{r})$ and 
$V^{\mathrm{nl}}(\mathbf{r},\mathbf{p})$  
are the local and nonlocal parts, 
 respectively.
The argument $(\mathbf{r},\mathbf{p})$
  is equivalente to the
explicit $(\mathbf{r},\mathbf{r}')$ nonlocal notation.\cite{ismailPRL01}
In case of
a local potential, i.e. $V= V(\mathbf{r})$, 
like that of all-electron schemes, 
we simply omit the contribution of $V^{\mathrm{nl}}(\mathbf{r},\mathbf{p})$
from the results that we have derived.

It is well known that the use of the LDA
leads to an underestimation of the band gap. A standard
procedure to correct for this is to
use the ``scissors approximation'', where the 
conduction bands are rigidly shifted in energy so that the band gap 
corresponds to the accepted 
experimental electronic band gap.\cite{levinePRL89,levinePRL91,delsolePRB93} 
This is often in fairly good agreement with the GW
band gap based on a more sophisticated calculation.\cite{hybertsenPRB86}
The LDA wave functions are used since they produce band
structures with dispersion relations similar to those predicted by the GW.
Mathematically, the scissors (non-local) operator 
$S$ is added to the unperturbed or unscissored Hamiltonian $H^{\mathrm{LDA}}_{0}$ ,
\begin{equation*}
H^\Sigma_{0}(\mathbf{r},\mathbf{p})=H^{\mathrm{LDA}}_{0}(\mathbf{r},\mathbf{p})+S(\mathbf{r},\mathbf{p})
\end{equation*}
where 
\begin{equation}
S(\mathbf{r},\mathbf{p})=\hbar \Delta\sum_{n}\int d^{3}k(1-f_{n})
|n\mathbf{k}\rangle\langle n\mathbf{k}|,
\label{hats}
\end{equation}
with $\hbar \Delta$  the rigid ($\mathbf{k}$-independent) energy correction to be
applied. 
The unscissored and scissored Hamiltonians satisfy 
\begin{align*}
H^{\mathrm{LDA}}_{0}(\mathbf{r},\mathbf{p})\psi _{n\mathbf{k}}(\mathbf{r}) &=
\hbar \omega^{\mathrm{LDA}}_{n}(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),\label{hamils} \\
H_{0}^\Sigma (\mathbf{r},\mathbf{p})\psi _{n\mathbf{k}}(\mathbf{r}) &=
\hbar \omega_{n}^\Sigma(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\end{align*}
where the scissor-shifted energies, 
$\omega_{n}^\Sigma(\mathbf{k})$ , are given by
\begin{equation*}
\omega_{n}^\Sigma(\mathbf{k})=\omega^\mathrm{LDA}_{n}(\mathbf{k})+(1-f_{n})\Delta.
\end{equation*}
We emphasize that the scissored and unscissored Hamiltonian 
have the same eigenfunctions, where
$\psi_{n\mathbf{k}}(\mathbf{r})=\langle\mathbf{r}|n\mathbf{k}\rangle=e^{i\mathbf{k}\cdot\mathbf{r}}u_{n\mathbf{k}}(\mathbf{r})$,
are the real space representations of the Bloch states $| n\mathbf{k}\rangle$ labelled 
by the band index $n$ and the crystal momentum $\mathbf{k}$, and $u_{n\mathbf{k}}(\mathbf{r})$
are cell periodic.

\subsection{Length Gauge Formalism}\label{lgf}

To calculate the optical response 
within the Length Gauge formalism, we take the 
interaction Hamiltonian as
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).
\label{rde}
\end{equation}
The 
treatment of the position operator
$\mathbf{r}$ for extended Bloch states, $\langle\mathbf{r}|n\mathbf{k}\rangle$ has 
been described in detail.\cite{adamsJCP53,blountSSP62} 
In particular, Ref.~\onlinecite{aversaPRB95} develops
the treatment of $\mathbf{r}$ for
the calculation of the nonlinear optical response.
$\mathbf{r}$ is split into the {\it intraband} 
($\mathbf{r}_i$) and {\it interband} ($\mathbf{r}_e$) parts, where 
$\mathbf{r}=\mathbf{r}_i+\mathbf{r}_e$. 
For $\mathbf{r}_e$ one uses
\begin{equation}
\langle n\mathbf{k} | \mathbf{r}_e | m\mathbf{k}'\rangle =
(1-\delta_{nm})\delta(\mathbf{k}-\mathbf{k}')\boldsymbol{\xi}_{nm}(\mathbf{k})
,
\label{rnminn}
\end{equation}
such that $\mathbf{r}_{e,nm}=0$ for $n=m$.
From Eq.~\eqref{mv} with $H_0\to
H^\Sigma_0$, we obtain
\begin{align}
\mathbf{r}_{e,nm}(\mathbf{k}) =
\boldsymbol{\xi}_{nm}(\mathbf{k})\equiv 
\mathbf{r}_{nm}(\mathbf{k}) 
&=
\frac{\mathbf{v}^\Sigma_{nm}(\mathbf{k})}{i\omega^\Sigma_{nm}(\mathbf{k})}
\quad\quad n\notin D_m 
,
\label{pmnrmn}
\end{align}  
where we defined
$\omega^\Sigma_{nm}(\mathbf{k})\equiv\omega^\Sigma_n(\mathbf{k})-\omega^\Sigma_m(\mathbf{k})$, and
$D_m$ are all the possible degenerate $m$-states. 
For the intraband part, $\mathbf{r}_i$ only appears in
commutators during the derivation of
the optical response. We use,\cite{aversaPRB95}
\begin{equation}
\langle n\mathbf{k} | [\mathbf{r}_i,{\cal O}] | m\mathbf{k}'\rangle
=i\delta(\mathbf{k}-\mathbf{k}')({\cal O}_{nm})_{;\mathbf{k}},
\label{conmri3n}
\end{equation}  
where
\begin{equation}
({\cal O}_{nm})_{;\mathbf{k}}=
\nabla_\mathbf{k} 
{\cal O}_{n m}(\mathbf{k}) 
- 
i 
{\cal O}_{nm}(\mathbf{k}) 
\left(
\boldsymbol{\xi}_{nn}(\mathbf{k}) 
-
\boldsymbol{\xi}_{mm}(\mathbf{k}) 
\right) 
,
\label{gendevnn}
\end{equation} 
is
the generalized derivative 
of the operator ${\cal O}$. 
The vectors $\boldsymbol{\xi}_{nn}(\mathbf{k})$ are defined in 
Ref.~\onlinecite{aversaPRB95} though they do not need to be 
calculated explicitly in what follows.

@@

Following Ref.~\onlinecite{sipePRB00} we take the
matrix elements of Eq.~\eqref{rhop} with the $H_{I}(t)$ of 
Eq.~\eqref{rde}, we obtain 
$(\tilde{\rho}^{(1)}(t))_{nm}=B_{nm}^{b}E^{b}(\omega)e^{i(\omega^\Sigma_{nm}-\tilde\omega)t}$,
with 
\begin{equation}
B_{nm}^{b}=\frac{e}{\hbar }\frac{f_{mn}r_{nm}^{b}}{\omega^\Sigma_{nm}-\tilde\omega},
\label{j.1}
\end{equation}
and 
\begin{equation}
(\tilde{\rho}^{(2)}(t))_{nm} = \frac{e}{i\hbar }\frac{1}{\omega^\Sigma_{nm}-2\tilde\omega}\bigg[
i\sum_{q }\Big(r_{nq }^{b}B_{q m}^{c}-B_{nq}^{c}r_{q m}^{b}
\Big)  
-(B_{nm}^{c})_{;k^{b}}\bigg]E^{b}(\omega)E^{c}(\omega)e^{i(\omega^\Sigma_{nm}-2\tilde\omega)t}
.
\label{j.2}
\end{equation}
We have used the fact that for a cold semiconductor $\partial
f_{n}/\partial \mathbf{k}=0$ and thus the intraband contribution to the linear
term vanishes identically. 
Note that the indices in Eq.~\eqref{j.2} are all different from each
other. This is due to the $f_{nm}$ factor in Eq.~\eqref{j.1}, 
and therefore $B^a_{nn}=0$. The dependence on $\mathbf{k}$ 
of all quantities is implicitly understood from 
this point forward.

\subsection{Layered Current Density}\label{cd}

>>>>

% @119: Introduction
% =====================================================
The most used framework for \textit{ab initio} calculations, Density 
Functional Theory (DFT) within the Local Density Approximation (LDA),
\cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
well understood that one has to include the many-body interaction to correct 
for this underestimation of the gap. In this context, the so-called GW 
approximation\cite{onidaRMP02} is known to correct the electronic gap of most 
semiconductors\cite{luceroJPCM12}. However this can be a very expensive
calculation and thus one uses the much simpler scissors operator scheme.
\cite{levinePRL89,levinePRL91,delsolePRB93}  
This allows us to ``open'' the DFT-LDA gap to 
its correct experimental or GW value for most bulk semiconductors.
%%%% Valerie and Nicolas remove these lines
% even when experiments are not available. 
%%%%
This approximation has already been used in linear optical calculations for 
surfaces, see \cite{kippPRL96}, thus improving the agreement with 
experimental results. 
In this context, to correct for the underestimation of the energy band gap of 
semiconductors Nastos et al.\cite{nastosPRB05} used the ``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to correctly include the 
many-body corrections through the scissors operator in the SH susceptibility.
Later, Cabellos \textit{et al}.\cite{cabellosPRB09} elaborated a derivation 
of the ``velocity gauge'' or ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the 
scissors operator and proved gauge invariance with respect to the length 
gauge. From these works it is clear the length gauge is a much better starting
point to obtain the surface SH (SSH) susceptibility, as will be elaborated 
in this article. However, these considerations are only valid for bulk 
semiconductors.

% =====================================================

% @1192: Half-slab vs. full-slab
% =====================================================

In Fig.~\ref{fig2}
we compare 
$\chi^{xxx}_{\mathrm{half-slab}}$  
vs. 
$\chi^{xxx}_{\mathrm{full-slab}}$ 
for the four different possibilities 
between including or not including the
effects of $\mathbf{v}^\mathrm{nl}$ or the scissors correction
$\hbar\Delta$.   
For these results we chose
$\hbar\Delta=0.5$ eV, that is the GW gap reported in
Refs.~\onlinecite{rohlfingPRB95,garciaCPC01}. 
This is justified by the fact that the surface states of the clean 
Si(001) surface investigated are found, from GW calculations 
\cite{rohlfingPRB95}, to shift rigidly and to keep their dispersion with 
respect to LDA.
We see that for all four instances the 
difference between responses is quite small.
Indeed, when the value $|\chi^{xxx}|$ 
is large the difference between the two is very small; 
when the value is small the difference increases only slightly, 
but the spectra is so close to zero that it is negligible. 
These differences would decrease as the number of atomic layers 
increases. We remark that 32 layers in the slab is more than enough 
to confirm that the extraction of the surface second-harmonic 
susceptibility from the $2\times 1$ surface is readily possible 
using the formalism contained in Eq.~\eqref{chis}.
We have confirmed that for the dihydride surface
$|\chi^{xxx}_{\mathrm{half-slab}}|\approx 0$ (not shown).
This confirms the validity of our theory and is the main result of
this article; through the proposed layer formalism we can calculate the surface SH
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$    
including
the
contribution of the nonlocal part of the pseudopotentials 
and the part of the many-body effects through the scissors correction.
Our scheme should work for any slab.

% =====================================================

% @409: Length Gauge Formalism
% =====================================================

According to ~\onlinecite{ismailPRL01}, we first start with the interaction 
Hamiltonian expressed in the velocity gauge, containing the nonlocal parts
$V^{nl}$ and $S$. Within the dipole approximation and using a gauge 
transformation, it can be transformed into an effective Hamiltonian 
\cite{Veniard, unpublished}
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).
\label{rde}
\end{equation}
The treatment of the position operator $\mathbf{r}$ for extended Bloch states 
is problematic and has been discussed in Refs.\cite{adamsJCP53,blountSSP62}
%%%% Valerie and Nicolas remove these lines
% In particular, Ref.~\onlinecite{aversaPRB95} develops the treatment of 
% $\mathbf{r}$ for the calculation of the nonlinear optical response.
%%%%
Following Ref.~\onlinecite{sipePRB00} we take the
matrix elements of Eq.~\eqref{rhop} with the $H_{I}(t)$ of 
Eq.~\eqref{rde}, we obtain 
$(\tilde{\rho}^{(1)}(t))_{nm}=B_{nm}^{b}E^{b}(\omega)e^{i(\omega^\Sigma_{nm}-\tilde\omega)t}$,
with 
\begin{equation}
B_{nm}^{b}=\frac{e}{\hbar }\frac{f_{mn}r_{nm}^{b}}{\omega^\Sigma_{nm}-\tilde\omega},
\label{j.1}
\end{equation}
and 
\begin{equation}
(\tilde{\rho}^{(2)}(t))_{nm} = \frac{e}{i\hbar }\frac{1}{\omega^\Sigma_{nm}-2\tilde\omega}\bigg[
i\sum_{q }\Big(r_{nq }^{b}B_{q m}^{c}-B_{nq}^{c}r_{q m}^{b}
\Big)  
-(B_{nm}^{c})_{;k^{b}}\bigg]E^{b}(\omega)E^{c}(\omega)e^{i(\omega^\Sigma_{nm}-2\tilde\omega)t}
.
\label{j.2}
\end{equation}
$\mathbf{r}$ is split into the {\it intraband} 
($\mathbf{r}_i$) and {\it interband} ($\mathbf{r}_e$) parts, where 
$\mathbf{r}=\mathbf{r}_i+\mathbf{r}_e$. 
For $\mathbf{r}_e$ one uses
\begin{equation}
\langle n\mathbf{k} | \mathbf{r}_e | m\mathbf{k}'\rangle =
(1-\delta_{nm})\delta(\mathbf{k}-\mathbf{k}')\boldsymbol{\xi}_{nm}(\mathbf{k})
,
\label{rnminn}
\end{equation}
such that $\mathbf{r}_{e,nm}=0$ for $n=m$.
From Eq.~\eqref{mv} with $H_0\to
H^\Sigma_0$, we obtain
\begin{align}
\mathbf{r}_{e,nm}(\mathbf{k}) =
\boldsymbol{\xi}_{nm}(\mathbf{k})\equiv 
\mathbf{r}_{nm}(\mathbf{k}) 
&=
\frac{\mathbf{v}^\Sigma_{nm}(\mathbf{k})}{i\omega^\Sigma_{nm}(\mathbf{k})}
\quad\quad n\notin D_m 
,
\label{pmnrmn}
\end{align}  
where we defined
$\omega^\Sigma_{nm}(\mathbf{k})\equiv\omega^\Sigma_n(\mathbf{k})-\omega^\Sigma_m(\mathbf{k})$, and
$D_m$ are all the possible degenerate $m$-states. 
For the intraband part, $\mathbf{r}_i$ only appears in
commutators during the derivation of
the optical response. We use,\cite{aversaPRB95}
\begin{equation}
\langle n\mathbf{k} | [\mathbf{r}_i,{\cal O}] | m\mathbf{k}'\rangle
=i\delta(\mathbf{k}-\mathbf{k}')({\cal O}_{nm})_{;\mathbf{k}},
\label{conmri3n}
\end{equation}  
where
\begin{equation}
({\cal O}_{nm})_{;\mathbf{k}}=
\nabla_\mathbf{k} 
{\cal O}_{n m}(\mathbf{k}) 
- 
i 
{\cal O}_{nm}(\mathbf{k}) 
\left(
\boldsymbol{\xi}_{nn}(\mathbf{k}) 
-
\boldsymbol{\xi}_{mm}(\mathbf{k}) 
\right) 
,
\label{gendevnn}
\end{equation} 
is
the generalized derivative 
of the operator ${\cal O}$. 
The vectors $\boldsymbol{\xi}_{nn}(\mathbf{k})$ are defined in 
Ref.~\onlinecite{aversaPRB95} though they do not need to be 
calculated explicitly in what follows.

@@

We have used the fact that for a cold semiconductor $\partial
f_{n}/\partial \mathbf{k}=0$ and thus the intraband contribution to the linear
term vanishes identically. 
Note that the indices in Eq.~\eqref{j.2} are all different from each
other. This is due to the $f_{nm}$ factor in Eq.~\eqref{j.1}, 
and therefore $B^a_{nn}=0$. The dependence on $\mathbf{k}$ 
of all quantities is implicitly understood from 
this point forward.

% =====================================================

% @1270: Results section C
% =====================================================
